---
title: "Get on your bike"
author: "Kuznik et al."
date: "2024-11-13"
output: pdf_document
---

# 1. Introduction 
The rapid urbanisation of recent decades has seen cities face challenges such as traffic congestion, air pollution, and limited parking spaces. In the face of these issues, rental bike systems have gained significant popularity, offering a cheap and eco-friendly alternative to more traditional commuting methods. Rental bikes provide a practical solution for short-distance travel, be it for students going to school, office workers heading to the workplace and even tourists exploring a city. In addition to promoting an active lifestyle, they eliminate the hassle of bike ownership, such as storage and maintenance. The growing demand of rental bike companies is reflected in the global market value of bike rentals, projected to reach a value of $11.3 billion in 2031, marking a 538% increase from the 2021 market value of $2.1 billion (Allied Market Research, 2021).
	The biggest challenge rental bike companies face is ensuring constant availability of bikes, as demand can fluctuate due to  numerous reasons. Bikes must be readily accessible to users at all times, making it essential for company leaders to be aware of peak hours, the impact of weather conditions and high-demand areas. In this project, we will use various machine learning models to predict the number of rented bikes required to meet customer demand under different circumstances, trying to figure out which of them performs best. The goal is to assist a fictional bike rental company operating in Seoul, South Korea, in planning the optimal size of their bike fleet.

# Dataset
To conduct our analysis, we will use a dataset called „Seoul Bike Sharing Demand“ which contains the count of public bicycles rented per hour in the Seoul Bike Sharing System with corresponding weather data and holiday information, collected between 2017 and 2018 (UCI Machine Learning Repository, 2020). We will enrich the dataset by adding columns that capture different times-related factors, such as whether a given day is a working day or a weekend. Additionally, we will categorise hours into specific periods to better capture daily trends in bike demand. 

# Structure
We will begin by examining the dataset and outlining the specific changes we made for the purpose of our analysis. Next, we will perform exploratory data analysis, focussing on correlations between variables to address potential multicollinearity issues. In the following chapters, we will fit different machine learning models to the data and use their results to predict the number of rental bikes needed in Seoul. The sequence of the models will be as follows: a linear model, a generalised linear model with a Poisson family, another with a Binomial family, a generalised additive model, a neural network and a support vector machine. Finally, we will compare these models to identify the best solution for our bike rental company.

```{r, echo = TRUE}


```

# 2. Data Transformation

## Libraries
```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(car)
library(multcomp)
```

The raw dataset was unsuitable for analysis and therefore required transformation. Seeing as the column names were unnecessarily complex and inconsistent, the first step was renaming them for simplicity. "Temperature.캜." was changed to "Temperature", "Humidity..." to "Humidity", "Visibility..10m." to "Visibility" and so forth. In the categorical variable "Functioning.Day", which indicated whether a given day was functioning for bike rentals, non-functioning days were removed as the number of rented bikes was obviously zero on those days. The column was subsequently dropped since its remaining values were identical, making it redundant.
  Next, time-related features were added to capture their potential impact on rented bikes. "Month" and "Weekday" columns were created to indicate the month and day of the week, followed by an "Is.Weekend" column to classify days as working days or weekends. To catch meaningful variations within a single day, the "Hours" column was grouped into segments such as morning rush hour, evening rush hour and late night. These segments were stored in a new variable "TimeOfDay". Finally, all categorical variables were converted to factor data types, completing the preparation of the dataset for analysis.

```{r}
# Changing the column names
colnames(data)[colnames(data) == "Temperature.캜."] <- "Temperature"
colnames(data)[colnames(data) == "Humidity..."] <- "Humidity"
colnames(data)[colnames(data) == "Dew.point.temperature.캜."] <- "Dew.Point.Temperature"
colnames(data)[colnames(data) == "Wind.speed..m.s."] <- "Wind.Speed"
colnames(data)[colnames(data) == "Solar.Radiation..MJ.m2."] <- "Solar.Radiation"
colnames(data)[colnames(data) == "Visibility..10m."] <- "Visibility"
colnames(data)[colnames(data) == "Rainfall.mm."] <- "Rainfall"
colnames(data)[colnames(data) == "Snowfall..cm."] <- "Snowfall"

# Removing non-functioning days
data <- data[data$Functioning.Day != "No", ]

# Adding Weekdays and weekend label
data <- data %>% 
  mutate(
    Date = as.Date(Date, format = "%d/%m/%Y"), # Changing Date format from character to date
    Weekday = weekdays(Date),  # Adding column with weekday
    Is.Weekend = ifelse(Weekday %in% c("Saturday", "Sunday"), "Yes", "No") # Determining whether a weekday is weekend
  )

#Adding Month Column
data$Month <- month(data$Date, label = TRUE, abbr = TRUE)

#Adding TimeofDay Column
data <- data %>%
    mutate(
    TimeOfDay = case_when(
      Hour >= 23 | Hour < 2 ~ "Early Night",
      Hour >= 2 & Hour < 7 ~ "Late Night",
      Hour >= 7 & Hour < 10 ~ "Morning Rush Hour",
      Hour >= 10 & Hour < 14 ~ "Mid-day",
      Hour >= 14 & Hour < 18 ~ "Afternoon",
      Hour >= 18 & Hour < 20 ~ "Evening Rush Hour",
      Hour >= 20 & Hour <= 22 ~ "Evening"
    )
  )

#Change datatypes for factor variables
data$Seasons <- as.factor(data$Seasons)
data$Holiday <- as.factor(data$Holiday)
data$Is.Weekend <- as.factor(data$Is.Weekend)
data$Weekday <- as.factor(data$Weekday)
data$TimeOfDay <- as.factor(data$TimeOfDay)
data$Hour <- as.factor(data$Hour)

```


# 3. Exploratory data analysis

```{r}
summary(data$Rented.Bike.Count)
med_rent <- median(data$Rented.Bike.Count)
med_rent

data$High.Rentals <- ifelse(data$Rented.Bike.Count > med_rent, 1, 0)
data$High.Rentals <- as.factor(data$High.Rentals)

# FACTOR VARIABLES
# Calculate proportions for each combination of Seasons and High.Rentals
data_prop <- data %>%
  group_by(Seasons, High.Rentals) %>%
  summarise(count = n()) %>%
  mutate(prop = count / sum(count)) %>%
  ungroup()

# Plot with percentages displayed on the bars
ggplot(data_prop, aes(x = Seasons, y = prop, fill = High.Rentals)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(aes(label = scales::percent(prop, accuracy = 1)), 
            position = position_fill(vjust = 0.5), size = 3) +
  labs(title = "Proportion of High Rentals by Season", x = "Season", y = "Proportion", fill = "High Rentals") +
  scale_fill_discrete(labels = c("No", "Yes")) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

# Calculate proportions for each combination of Is.Weekend and High.Rentals
data_prop <- data %>%
  group_by(Month, High.Rentals) %>%
  summarise(count = n()) %>%
  mutate(prop = count / sum(count)) %>%
  ungroup()

# Plot with percentages displayed on the bars
ggplot(data_prop, aes(x = Month, y = prop, fill = High.Rentals)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(aes(label = scales::percent(prop, accuracy = 1)), 
            position = position_fill(vjust = 0.5), size = 3) +
  labs(title = "Proportion of High Rentals by Months", x = "Months", y = "Proportion", fill = "High Rentals") +
  scale_fill_discrete(labels = c("No", "Yes")) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

# Calculate proportions for each combination of Is.Weekend and High.Rentals
data_prop <- data %>%
  group_by(Is.Weekend, High.Rentals) %>%
  summarise(count = n()) %>%
  mutate(prop = count / sum(count)) %>%
  ungroup()

# Plot with percentages displayed on the bars
ggplot(data_prop, aes(x = Is.Weekend, y = prop, fill = High.Rentals)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(aes(label = scales::percent(prop, accuracy = 1)), 
            position = position_fill(vjust = 0.5), size = 3) +
  labs(title = "Proportion of High Rentals by Weekend", x = "Weekend", y = "Proportion", fill = "High Rentals") +
  scale_fill_discrete(labels = c("No", "Yes")) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

# Calculate proportions for each combination of Is.Weekend and High.Rentals
data_prop <- data %>%
  group_by(Weekday, High.Rentals) %>%
  summarise(count = n()) %>%
  mutate(prop = count / sum(count)) %>%
  ungroup()

# Plot with percentages displayed on the bars
ggplot(data_prop, aes(x = Weekday, y = prop, fill = High.Rentals)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(aes(label = scales::percent(prop, accuracy = 1)), 
            position = position_fill(vjust = 0.5), size = 3) +
  labs(title = "Proportion of High Rentals by Weekday", x = "Weekday", y = "Proportion", fill = "High Rentals") +
  scale_fill_discrete(labels = c("No", "Yes")) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

# Calculate proportions for each combination of Is.Weekend and High.Rentals
data_prop <- data %>%
  group_by(TimeOfDay, High.Rentals) %>%
  summarise(count = n()) %>%
  mutate(prop = count / sum(count)) %>%
  ungroup()

# Plot with percentages displayed on the bars
ggplot(data_prop, aes(x = TimeOfDay, y = prop, fill = High.Rentals)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(aes(label = scales::percent(prop, accuracy = 1)), 
            position = position_fill(vjust = 0.5), size = 3) +
  labs(title = "Proportion of High Rentals by Time of Day", x = "TimeOfDay", y = "Proportion", fill = "High Rentals") +
  scale_fill_discrete(labels = c("No", "Yes")) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

# Calculate proportions for each combination of Is.Weekend and High.Rentals
data_prop <- data %>%
  group_by(Hour, High.Rentals) %>%
  summarise(count = n()) %>%
  mutate(prop = count / sum(count)) %>%
  ungroup()

# Plot with percentages displayed on the bars
ggplot(data_prop, aes(x = Hour, y = prop, fill = High.Rentals)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(aes(label = scales::percent(prop, accuracy = 1)), 
            position = position_fill(vjust = 0.5), size = 3) +
  labs(title = "Proportion of High Rentals by Hour", x = "TimeOfDay2", y = "Proportion", fill = "High Rentals") +
  scale_fill_discrete(labels = c("No", "Yes")) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()


# CONTINOUS VARIABLES
# Summary statistics for continuous variables by High Rentals
data %>%
  group_by(High.Rentals) %>%
  summarise(
    avg_temperature = mean(Temperature, na.rm = TRUE),
    avg_humidity = mean(Humidity, na.rm = TRUE),
    avg_wind_speed = mean(Wind.Speed, na.rm = TRUE),
    avg_visibility = mean(Visibility, na.rm = TRUE)
  )

# Box plot for Temperature by High Rentals
ggplot(data, aes(x = High.Rentals, y = Temperature)) +
  geom_boxplot() +
  labs(title = "Temperature vs. High Rentals", x = "High Rentals", y = "Temperature (°C)") +
  scale_x_discrete(labels = c("0" = "Low", "1" = "High")) +
  theme_minimal()

# Density plot for Temperature
ggplot(data, aes(x = Temperature, fill = High.Rentals)) +
  geom_density(alpha = 0.5) +
  labs(title = "Temparature for High and Low Rentals", x = "Temperature (°C)", y = "Density") +
  scale_fill_manual(name = "High Rentals", values = c("0" = "red", "1" = "blue")) +
  theme_minimal()

# Box plot for Humiditiy by High Rentals
ggplot(data, aes(x = High.Rentals, y = Humidity)) +
  geom_boxplot() +
  labs(title = "Humidity vs. High Rentals", x = "High Rentals", y = "Humidity (%)") +
  scale_x_discrete(labels = c("0" = "Low", "1" = "High")) +
  theme_minimal()

# Density plot for Humidity
ggplot(data, aes(x = Humidity, fill = High.Rentals)) +
  geom_density(alpha = 0.5) +
  labs(title = "Humidity Distribution for High and Low Rentals", x = "Humidity (%)", y = "Density") +
  scale_fill_manual(name = "High Rentals", values = c("0" = "red", "1" = "blue")) +
  theme_minimal()
```


# 4. Modeling 
##    a) Linear Model
Note: We want to predict the number of bikes rented in Seoul, looking at several different variables. Logically, our response variable should be Rented.Bike.Count. However, as that is a count variable, it is not ideal for a standard linear model. As the only two continuous variables in our data set measure temperatures, fitting a linear model with either as the response variable would not be valuable to our analysis. Thus, we will transform Rented.Bike.Count with the square-root function to make it more suitable as a response variable for linear regression.
  As season and month are highly correlated, month is excluded in the model. Similarly, because Is.Weekend and Weekday are strongly correlated, Weekday is excluded in the model. The same goes for temperature and dew.point.temperature, dew.point.temperature is excluded in the model.

```{r}
lm.bikes <- lm(sqrt(Rented.Bike.Count) ~ Temperature + 
                Humidity + Wind.Speed + Visibility + 
                Solar.Radiation + Rainfall + Snowfall + 
                Seasons + Holiday + Is.Weekend + TimeOfDay, 
              data=data)
summary(lm.bikes)
```

To check the effects of categorical variables on the response variable, we use the drop1 function.

```{r}
drop1(lm.bikes, test = 'F')
```
Since there are some variables who don't seem to have any effect on the rental bike count, we will check if we can fit a better model.In the second version of the model, we remove the variables Wind.Speed, Visibility, Solar.Radiation and Snowfall to see how the model performance changes.

```{r}
lm.bikes.2 <- update(lm.bikes, .~. - Wind.Speed - Visibility - Solar.Radiation - Snowfall)
summary(lm.bikes.2)
```

Like with the first version we use the drop1 function to check the effects of categorical variables on the rental bike count.

```{r}
drop1(lm.bikes.2, test = 'F')
```

Let's check the model for multicollinearity

```{r}
vif(lm.bikes.2)
```

In the second model, the GVIF values of Temperature and Seasons are just below the threshold of 5 so they should be fine to keep in the model. Let's now compare the two models to see which of them performs better in predicting the rental bike count in Seoul. We look at the adjusted R-squared of both and see that they are almost identical at around 0.704, suggesting that using the simpler model is justified. Next we check the fit of the models.

```{r}
AIC(lm.bikes, lm.bikes.2)
BIC(lm.bikes, lm.bikes.2)
```

The AIC values are very similar but the second model has a lower BIC value, indicating that the simpler model is a better fit. Finally, we compare the models with an F-test.

```{r}
anova(lm.bikes, lm.bikes.2)
```

There is no significant difference between the two models, which means we can opt for the simpler one. With the second model chosen for the analysis, let's interpret the coefficients.

```{r}
summary(lm.bikes.2)
drop1(lm.bikes.2, test= 'F')
```

As the model was fitted to the square root of rented bikes, the coefficients of predictor variables estimate the change in the square root of rented bikes, not the direct number of bikes rented. To interpret the effect on the actual number of bikes rented, we need to square the predicted value (sum of the intercept and the coefficient). The difference between this squared value and the square of the intercept gives the predicted change in rented bikes due to a variable.

#Intercept:
There is strong evidence that the mean number of rental bikes on a normal working day in autumn during the morning rush hour is not zero. Around 1267 bikes are expected to be rented in this situation.

#Temperature:
There is strong evidence that for each increase in temperature by one degree, while all other variables are kept constant, an extra 34 bikes will be rented.

#Humidity:
There is strong evidence that for each increase in humidity by one percentage point, while all other variables are kept constant, 11 fewer bikes will be rented.


#Rainfall:
There is strong evidence that for each increase in rainfall by 1 mm, while all other variables are kept constant, 117 fewer bikes will be rented.

We now interpret the coefficients of categorical variables, it is important to note that the observed values are all in comparison to the reference level of each variable but not between each other. In the next section we will compare the categorical variables Seasons and TimeOfDay between each level.

#Seasons:
There is strong evidence that for a spring day where other variables are kept constant, 226 fewer bikes will be rented than in autumn. For a summer day, a decrease of 221 bikes is expected and for a winter day, a big decrease of 502 bikes is expected compared to autumn. Autumn coming out on top compared to the other seasons might be down to reasons like mild weather often associated with autumn and people getting back into rhythm after the summer holidays.

#Holiday:
There is strong evidence that on a public holiday, while all other variables are kept constant, 203 fewer bikes will be rented than on a normal day. This suggests that people use the rental bikes more as a means of transport, i.e. to get to and from school or work, rather than for leisure when they have free time.

#Is.Weekend:
There is strong evidence that on a weekend, while all other variables are kept constant, 113 fewer bikes will be rented. Similar to public holidays, weekends where more people have free time don't seem to increase people's needs for rental bikes.

#TimeOfDay:
There is strong evidence that during mid-day, if all other variables are kept constant, 384 fewer bikes will be rented compared to the morning rush hour. Similarly, in the afternoon, a decrease of 217 rented bikes is expected. In the evening rush hour, an increase of 364 bikes is expected compared to the morning rush hour and in the evening, there is an increase of 146 bikes. In the early and late night, the decrease compared to the morning rush hour is 282 and 681 bikes respectively. The busiest times for rented bikes seem to be the two rush hours and the evening. Between the rush hours there is a dip in the rental activity and unsurprisingly, during the night it plummets.

We now check the difference between all levels of seasons and time of day.

```{r}
seasonal.differences <- glht(lm.bikes.2,
                     linfct = mcp(Seasons = "Tukey"))
summary(seasonal.differences)

ToD.differences <- glht(lm.bikes.2,
                     linfct = mcp(TimeOfDay = "Tukey"))
summary(ToD.differences)
```

There is strong evidence that no two levels of the time of day variable have the same effect on the number of rented bikes. For seasons, the same situation is seen except that there is no evidence that summer and spring have different effects on the number of rented bikes.

Valtýr Már Michaelsson took the lead on the linear model.

    i) Model implementation
    ii) Result interpretation
    ii) Cross Validation*

  b) Generalised Linear Model with family set to Poisson
    i) Model implementation
    ii) Result interpretation
    ii) Cross Validation*
    
  c) Generalised Linear Model with family set to Binomial
    i) Model implementation
    ii) Result interpretation
    ii) Cross Validation*
    
  d) Generalised Additive Model
    i) Model implementation
    ii) Result interpretation
    ii) Cross Validation*

  e) Support Vector Machine (SVM)
    i) Model implementation
    ii) Result interpretation
    ii) Cross Validation*

  f) Neural Network
    i) Model implementation
    ii) Result interpretation
    ii) Cross Validation*

5. Usage of AI

6. Conclusion
